Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-06-14 MS/ms.Rmd
I think we need to 1) add another section on the apples and oranges problem to make it super clear to the reader what exactly it is. The introduction may not currently achieve this well enough, and it seems like an abrupt transition to survey from the introduction. 2) Seperate out goasl of survey before jumping into the methods.

* Corespondance:

@@

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Distilling published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance) in a meta-analysis, and compared across broad sets of research designs, study systems and species is its core objective and strength. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing “apples with oranges”. Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-analytic models to both account for factors driving differences across studies and make them more comparable. In addition, we derive new effect size measures that provide comparative physiologists with a means to directly make effect sizes comparable without the need to resort to more complex statistical models that may be harder to interpret. Our ‘new’ effect sizes and corresponding sampling variances help physiologists deal with common treatment differeces across studies; allowing researchers to compare both mean differences (e.g., $Q_{10}$ to compare mean differences in physiological rates over 10 degrees) and associated differences in variance (e.g, $S_{Q10}$ for comparing variability in in physiological rates over 10 degrees) for physiological traits. The new effect sizes we propose, in combination with existing meta-analytic models, will pave the way for comparative physiologists to explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

The goal of our review is to briefly overview meta-analytic practices and reassure comparative physiologists that heterogeneity among effect sizes can be comfortably dealt with by modifying existing effect sizes measures and using multilevel meta-regression models. We first conduct a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic approaches - assessing what types of effect sizes are commonly used, how often effect size estimates are impacted by continuous treatment differences and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We then introduce new effect sizes and their assocaited sampling variance to provide comparative physiologists opportunities to explictly incorperate treatment differences into the effect size estimates to easy their interpretation. We then describe how more complex treatment differences, such as dosage differences, which may have more complex non-linear effects, can be accomodated using multilevel meta-regression models. Expanding the meta-analtic toolkit for comparative physiologists will provide new opportunities to address global scale problems that attempt to undersatnd how organisms will cope with rapidly changing environments in the future.

## **Survey of Meta-Analytic Practices in Comparative Physiology**
#### *Literature Search and Survey Data Collection*
To understand meta-analytic practices in comparative physiology, we searched for meta-analyses and comparative analyses that tackled questions that relied on using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptable to variation in how treatments are applied across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored and if so it was usable (i.e., in a format like .txt).  All searches were conducted in Scopus between May 16 - 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macrophysiology' in their title and abstract, restricting our search to key comparative physiology journals (deatiled in Table X - Supplementary Materials) within the last 6 years. We expected this to provide a contemporay overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We awknowledge that this may have missed some important, and influential, quantitative synthese in the field of comparative physiology, but likely provided a random sample to inform on the questions we sought to address (see below).

>>>>

@@

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Distilling published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance) in a meta-analysis, and compared across broad sets of research designs, study systems and species is its core objective and strength. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing “apples with oranges”. Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-analytic models to both account for factors driving differences across studies and make them more comparable. In addition, we derive new effect size measures that provide comparative physiologists with a means to directly make effect measures comparable without the need to resort to more complex statistical models that may be harder to interpret. Our ‘new’ effect sizes and corresponding sampling variances help physiologists deal with common treatment differeces across studies; allowing researchers to compare both mean differences and associated differences in variance for physiological traits. The new effect sizes we propose, in combination with existing meta-analytic models, will pave the way for comparative physiologists to explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

The goal of our review is to briefly overview meta-analytic practices and reassure comparative physiologists that heterogeneity among effect sizes can be comfortably dealt with by modifying existing effect sizes measures and using multilevel meta-regression models. We first conduct a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic approaches - assessing what types of effect sizes are commonly used, how often effect size estimates are impacted by continuous treatment differences and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We then introduce new effect sizes and their assocaited sampling variance to provide comparative physiologists opportunities to explictly incorperate treatment differences into the effect size estimates to ease their interpretation. We then describe how more complex treatment differences, such as dosage differences, which may have non-linear effects, can be accomodated using multilevel meta-regression models. We hope to show how expanding the meta-analtic toolkit for comparative physiologists will provide new opportunities to address global scale problems that attempt to understand how organisms will cope with rapidly changing environments in the future.

## **The Apples and Oranges 'Problem': How lack of comparability can make effects vary widely across studies**
 I'm thinking now, in re-reading this, that we could use a slightly expanded version of paragraph 4 taht describes this problem in a little more detail. This is likely the place for a figure, which I am having some trouble envisaging where one is best placed. It just otherwise seems like a bit of a jump to otherwise head straight into the survey

## **Survey of Meta-Analytic Practices in Comparative Physiology**
As discussed above, it is clear that different types of questions and experimental approaches will be more or less susceptable to the 'apples and oranges' problem. As such, before embarking on our path to 'meta-analytic enlightenment', we conducted a literatrure survey to understand how widespread the problem may be in comparative physiology and gauge the ways in which it is currently being addressed, if at all. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptable to variation in how treatments are applied across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

#### *Literature Search and Survey Data Collection*
All searches were conducted in Scopus between May 16 - 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macrophysiology' in their title and abstract, restricting our search to key comparative physiology journals (deatiled in Table X - Supplementary Materials) within the last 6 years. We expected this to provide a contemporay overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We awknowledge that this may have missed some important, and influential, quantitative synthese in the field of comparative physiology, but likely provided a random sample to inform on the questions we sought to address (see below).
